An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects in Time-Frequency Representations

### 2020-07-10 Notes/BoundEffRed.tex
code

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the boundary effects reduction on TF representations to achieve a real-time signal analysis. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient for the extension of signals formed by a sum of sine waves. Moreover, the numerical results show that this strategy allows us to switch to a real-time implementation of {\sf BoundEffRed} is possible, due to the low computation time, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many time-frequency representations. 
%Other numerical examples are also included, such as the application of {\sf BoundEffRed} to other types of biomedical signals such as ECGs.

Various improvements can be considered to make the algorithm more robust to certain situations. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting is no longer relevant, and the TF representations are not ideal. A preliminary step should then be added to the algorithm to select the portions without sufficient quality, in order to reject the portions for which a better forecasting scheme is needed. A more fundamental improvement would be to perform the prediction step directly in the TF plane, or even with the aim of specifically predicting the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.

>>>>

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the reduction of boundary effects on TF representations in real-time. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient for the extension of signals formed by a sum of sine waves. Moreover, the numerical results show that this strategy allows us to switch to a real-time implementation of {\sf BoundEffRed} is possible, due to the low computation time, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many time-frequency representations. 
%Other numerical examples are also included, such as the application of {\sf BoundEffRed} to other types of biomedical signals such as ECGs.

Various improvements can be considered to make the algorithm more robust. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting is no longer relevant, and the TF representations are not ideal. A preliminary step should then be added to the algorithm to detect the portions to be rejected, for which a better forecasting scheme is needed. A more fundamental improvement would be to perform the prediction step directly in the TF plane, or even with the aim of specifically predicting the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.

### 2020-07-10 Notes/results.tex
code

\paragraph{Summary}
Both previous experimental results combined with the theoretical asymptotic equation~\eqref{eq:cov.error.2} allow us to describe the influence of the noise variance and the size of the training dataset on the variance of the forecasting noise, which is empirically summarized as follows:
\begin{equation}
\gamma[N-1+\ell,N-1+\ell] \underset{K\to\infty}{\approx} \dfrac{\sigma^2}{K}g[\ell] .
\end{equation} 
where $g$ is a bounded positive function. The empirical result is coherent with the theoretical result provided by Theorem~\ref{th:error}.

@@

\subsubsection{Adaptive harmonic model}
\label{ssse:res.ahm}
We now consider a signal satisfying the AHM so that the instantaneous frequencies and amplitudes of its components vary over time. The deterministic component $\bz$ of the random vector $\bx$ (constructed following the model~\eqref{eq:model.noise}) takes the following form, for all $n\in\{1,\ldots,N\}$:
\[
\bx[n] = \cos\left(2\pi \phi_1[n] \right) + R[n]\cos\left(2\pi \phi_2[n] \right) \ ,
\] 
where the instantaneous amplitude $R$ is given by:
\[
R[n] = 1.4 + 0.2\cos\left(4\pi\frac{n}{N}\right)\ ,
\]
and the instantaneous phases are such that:
\begin{align*}
\phi_1[n] &= \frac{p_1}{M}\left( n + \frac{0.01}{2\pi}\cos\left(2\pi\frac{n}{N}\right) \right) \\[-1mm]
\phi_2[n] & = p_2\frac{n}{M} + \frac{20}{2N\fs}n^2
\end{align*}
Besides, the additive noise is chosen to be Gaussian: $\bw\sim\cN(\bzero,\bI)$. Numerically, we take: $N=10^4$, $M=750$, $p_1=10$, $p_2=23$.

@@

\subsubsection{Metrics}
The quality of the boundary effects reduction is evaluated directly on the TF representation. To that aim, we compare the obtained representation to the optimal representation $\ccF_N^\mathrm{opt}(\bx)$, defined as the restriction of the representation of the ground-truth extended signal $\bx^\mathrm{ext}$. Therefore, we have:
\begin{equation}
\ccF_N^\mathrm{opt}(\bx) = \cR\left( \ccF_{N+2L}(\bx^\mathrm{ext}) \right) \ .
\label{eq:opt.TFR}
\end{equation}

In the aim of comparing the different techniques, we use a criterion, proposed in~\cite{Daubechies16conceft}, that quantify the distance between a given TF representation and the optimal one. It is built in analogy with the optimal transport distance, which enables quantifying the distance between two probability density functions. Let us generically denote a time frequency representation $\ccQ$. Then, for $t$ fixed, we consider the following probability density function: $p_\ccQ^t(\xi) = \left. |\ccQ(\xi,t)|^2 \middle/ \int_\RR |\ccQ(\nu,t)|^2\dd\nu \right.$. At each instant $t$, we can then determine the optimal transport distance $d_{t}$ between the two densities. It is given by the $L^1$ norm of the difference between the associated distribution functions. In other words, we have:
\begin{equation*}
d_{t}(\ccQ,\ccF_0) = \int_\RR\left|\tilde P_{\ccQ}^t(\xi)-  P_{\ccF_0}^t(\xi)\right|\dd\xi\ ,
\end{equation*}
where $P_{\ccQ}^t(\xi)=\int_{-\infty}^\xi p_{\ccQ}^t(\nu)\dd\nu$ and $\tilde P_{\ccF_0}^t(\xi)=\int_{-\infty}^\xi\tilde p_{\ccF_0}^t(\nu)\dd\nu$. The \textit{optimal transport distance} quantifies the proximity between the estimated and actual instantaneous frequencies while favoring the sparsity of the estimated TF representation. That is why, the performance index $D(\ccQ)$ for the reduction of boundary effects of a given TF representation $\ccQ$ is given by the averaged ratio between the optimal transport distance to the optimal TF representation~\eqref{eq:opt.TFR} and the optimal transport distance of to the optimal TF representation, that is:
\begin{equation}
D(\ccQ) = \dfrac1{|I|}\bigintss_I\ \frac{d_{t}\left(\ccQ,\ccF_N^\mathrm{opt}\right)}{d_{t}\left(\ccF_N,\ccF_N^\mathrm{opt}\right)}\ \dd t\ .
\label{eq:index.perf}
\end{equation}
Thus, $D(\ccQ)<1$ means a reduction of the boundary effects. Let us evaluate the quality of the boundary effects reduction on biomedical signals.

\subsubsection{Respiratory signal}
We first consider a 6 hours 20 minutes-long respiratory signal. This signal is sampled at $\fs=100$~Hz. A zoom on a small portion of the signal is displayed in Fig.~\ref{fig:tho}.

>>>>

\paragraph{Summary}
Both previous experimental results combined with the theoretical asymptotic equation~\eqref{eq:cov.error.2} allow us to describe the influence of the noise variance and the size of the training dataset on the variance of the forecasting noise, empirically summarized as follows:
\begin{equation}
\gamma[N-1+\ell,N-1+\ell] \underset{K\to\infty}{\approx} \dfrac{\sigma^2}{K}g[\ell] .
\end{equation} 
where $g$ is a bounded positive function. The empirical result is coherent with the theoretical result provided by Theorem~\ref{th:error}.

@@

\subsubsection{Adaptive harmonic model}
\label{ssse:res.ahm}
We now consider a signal satisfying the AHM so that the instantaneous frequencies and amplitudes of its components vary over time. The deterministic component $\bz$ of the random vector $\bx$ (constructed following the model~\eqref{eq:model.noise}) takes the following form, for all $n\in\{1,\ldots,N\}$:
\[
\bx[n] = \cos\left(2\pi \phi_1[n] \right) + R[n]\cos\left(2\pi \phi_2[n] \right) \ ,
\] 
where the instantaneous amplitude $R$ is given by:
\[
R[n] = 1.4 + 0.2\cos\left(4\pi\frac{n}{N}\right)\ ,
\]
and the instantaneous phases are such that:
\begin{align*}
\phi_1[n] &= \frac{p_1}{M}\left( n + \frac{0.01}{2\pi}\cos\left(2\pi\frac{n}{N}\right) \right) \\[-1mm]
\phi_2[n] & = p_2\frac{n}{M} + \frac{20}{2N\fs}n^2
\end{align*}
Besides, the noise is chosen to be Gaussian: $\bw\sim\cN(\bzero,\bI)$, and we take: $N=10^4$, $M=750$, $p_1=10$, $p_2=23$.

@@

\subsubsection{Metrics}
The quality of the boundary effects reduction is evaluated directly on the TF representation. To that aim, we compare the obtained representation to the optimal representation $\ccF_N^\mathrm{opt}(\bx)$, defined as the restriction of the representation of the ground-truth extended signal $\bx^\mathrm{ext}$. Therefore, we have:
\begin{equation}
\ccF^\mathrm{opt}(\bx_N) = \cR\left( \ccF(\bx^\mathrm{ext}) \right) \ .
\label{eq:opt.TFR}
\end{equation}

In the aim of comparing the different techniques, we use a criterion, proposed in~\cite{Daubechies16conceft}, that quantify the distance between a given TF representation and the optimal one. It is built in analogy with the optimal transport distance, which enables quantifying the distance between two probability density functions. Let us generically denote a time frequency representation $\ccQ$. Then, for $t$ fixed, we consider the following probability density function: $p_\ccQ^t(\xi) = \left. |\ccQ(\xi,t)|^2 \middle/ \int_\RR |\ccQ(\nu,t)|^2\dd\nu \right.$. At each instant $t$, we can then determine the optimal transport distance $d_{t}$ between the two densities. It is given by the $L^1$ norm of the difference between the associated distribution functions. In other words, we have:
\begin{equation*}
d_{t}(\ccQ,\ccF_0) = \int_\RR\left|\tilde P_{\ccQ}^t(\xi)-  P_{\ccF_0}^t(\xi)\right|\dd\xi\ ,
\end{equation*}
where $P_{\ccQ}^t(\xi)=\int_{-\infty}^\xi p_{\ccQ}^t(\nu)\dd\nu$ and $\tilde P_{\ccF_0}^t(\xi)=\int_{-\infty}^\xi\tilde p_{\ccF_0}^t(\nu)\dd\nu$. The \textit{optimal transport distance} quantifies the proximity between the estimated and actual instantaneous frequencies while favoring the sparsity of the estimated TF representation. That is why, the performance index $D(\ccQ)$ for the reduction of boundary effects of a given TF representation $\ccQ$ is given by the averaged ratio between the optimal transport distance to the optimal TF representation~\eqref{eq:opt.TFR} and the optimal transport distance of to the optimal TF representation, that is:
\begin{equation}
D(\ccQ) = \dfrac1{|I|}\bigintss_I\ \frac{d_{t}\left(\ccQ,\ccF^\mathrm{opt}\right)}{d_{t}\left(\ccF,\ccF^\mathrm{opt}\right)}\ \dd t\ .
\label{eq:index.perf}
\end{equation}
Thus, $D(\ccQ)<1$ means a reduction of the boundary effects. Let us evaluate the quality of the boundary effects reduction on biomedical signals.

\subsubsection{Respiratory signal}
We first consider a respiratory signal of 6 hours 20 minutes. This signal, a small portion of which is displayed in Fig.~\ref{fig:tho}, is sampled at $\fs=100$~Hz.
